---
title: "2025 Final"
subtitle: "https://github.com/bonetati/ENVS-193DS_spring-2025_final"
output: html_document
---

```{r setup, message=FALSE}
library(tidyverse)
library(here)
library(DHARMa)
library(MuMIn)

sst <- read_csv(here("data", "SST_update2023.csv"))
occdist <- read_csv("data/occdist.csv")
```

1a. 
In part 1, they used a pearson correlation test to compare the relationship between distance from head water and annual total nitrogen load. In part 2, they used a one-way ANOVA to test differences in mean annual nitrogen load between five different sources, urban land, atmospheric deposition, fertilizer, waste water treatment, and grasslands.

1b. To learn more information we could do a post-hoc pairwise comparison test which would identify which of the specific sources differ significantly from each other in average nitrogen load. The second piece of information that would be useful to to have is the group means and standard deviations for each nitrogen source since this would help better understand the differences in nitrogen load between sources. This is important as it provides biological context.

1c.
Part 1: We found that total nitrogen load tends to increase with distance from headwater, suggesting that agricultural runoff accumulates downstream. Using the Pearson correlation test were r = correlation coefficient, p = 0.03, and α = significance level. 
Part 2: We found that average nitrogen load varied significantly among sources, with fertilizer and wastewater treatment appearing to contirbute more nitrogen loads than grasslands or atmospheric deposition. Using the statistical test of a one-way ANOVA were F = test statistic, df = degrees of freedom, p = 0.02, α = significance level.

2a.
```{r}
# Clean/summarize the data
sst_clean <- sst %>%
  # Convert date column to Date object
  mutate(date = ymd(date), 
         # Extract year,convert to a factor
         year = factor(year(date)), 
          # Extract the month, abbreviate and make ordered factor
         month = factor(month(date, label = TRUE, abbr = TRUE), 
                        levels = month.abb, ordered = TRUE)) %>%
  # Group by year and month
  group_by(year, month) %>%
  # Mean sea surface temperature for each group
  summarize(mean_monthly_sst = mean(temp, na.rm = TRUE), .groups = "drop")

#Display 5 random rows
sst_clean %>%
  slice_sample(n = 5)

#Show structure
str(sst_clean)
```

2b.
```{r}

#Filter only 2018-2023
sst_subset <- sst_clean %>%
  filter(year %in% c("2018", "2019", "2020", "2021", "2022", "2023"))

#Graph
ggplot(sst_subset, aes(x = month, y = mean_monthly_sst, group = year, color = year)) +
  #Line and point geometry
  geom_line(linewidth = 0.5) +
  geom_point(size = 2) +
  # Axis labels
  labs(
    x = "Month",
    y = "Mean monthly sea surface temperature (°C)",
    color = "Year"
  ) +
  #Gradient from light blue to dark blue
  scale_color_manual(
    values = colorRampPalette(c("lightblue", "darkblue"))(6)
  ) +
  
  #Legend inside the panel + theme
  theme_classic() +
  theme(
    legend.position = c(0.15, 0.7),  
  )
```


3a.
The response variable is Swift Parrot nest box occupancy, where 1 indicates that a nest box was occupied by a Swift Parrot and 0 indicates that it was not occupied. 

3b.
Swift Parrots are a critically endangered species, while Common Starlings and Tree Martins are not. The study compares these species to determine whether artificial nest boxes are successful in supporting the conservation of Swift Parrots or unintentionally favoring the non-endagered species.

3c.
The two seasons refer to data collected in 2016 and 2019. In 2016, Swift Parrots were known to be breeding in the area of study, while in 2019, they were not. This difference changes the availability and competition for nest boxes.

3d.
| Model number | Season                 | Distance to forest edge | Model description                   |
| ------------ | ---------------------- | ----------------------- | ----------------------------------- |
| 1            | Not included           | Not included            | Null model                          |
| 2            | Included               | Included                | Saturated model (season + distance) |
| 3            | Included               | Not included            | Season only model                   |
| 4            | Not included           | Included                | Distance only model                 |

3e.
```{r}

#Clean data
 occdist <- occdist|>
  mutate(
    season = factor(season),
    sp = as.numeric(sp)
  )

#Run models
mod1 <- glm(sp ~ 1, data = occdist, family = binomial) #Null model
mod2 <- glm(sp ~ season + `edge distance`, data = occdist, family = binomial) #Saturated model
mod3 <- glm(sp ~ season, data = occdist, family = binomial) #Season only
mod4 <- glm(sp ~ `edge distance`, data = occdist, family = binomial) #Distance only
```









